Is malaria control profitable? Return on investment of privately-managed residential fumigations at a large sugarcane processing facility in Southern Mozambique

### 2018-04-20 inst/rmd/01_introduction.Rmd
major overhaul

From the persepctive of the private sector, however, investing in malaria control is not so clear-cut. Public health interventions targetting malaria - and their corresponding cost-effectiveness evaluations - most often focus on impacts pertaining to public welfare, such as an increase in life years adjusted for disability or quality  [@Goodman1999] [@Shretta2016] [@Lee2017] [@Hanson2004]. Though population-level health is certainly of importance to businesses, and improvements in health incidentally improve the economy at all levels [@World1999] [@Bloom2008] [@Vecchi_2013], these improvements may be too disperse or long-term to incentivize private sector involvement in health campaigns.

@@

A significant sector of the economy in Mozambique is dominated by large-scale foreign direct investment projects [@Robbins2012], and the role of the private sector in health generally, and malaria specifically, is unequivocally important. Large agriculture and extractive industry firms take up wide swaths of land and employ hundreds of thousands  [@German2013]. The Mozambican state has encouraged large-scale entreprise with the aim of general economic development [@Buur2012]. And where large firms exist, they often take on social roles such as housing and health care [@Winkler]. At times, this role is necessary from a purely practical standpoint; in other cases, it is employed under the auspices of "corporate social responsibility" [@Azemar2009]. Regardless of the language used, it is clear that private industry plays an important role in public health in Mozambique [@Robbins2012] [@CastelBranco2014].

Several cases exist of foreign firms engaging in large-scale malaria control campaigns [@Mouzin2011] [@Han] [@Bennett_2017] [@Kaula_2017]. But these studies generally consider population health as the outcome measure of interest, rather than worker absenteeism or productivity. Similarly, they often neglect to differentiate between those clinical costs which are absorbed by the local health system versus those which are absorbed by the firm itself. In the literature, making the "investment case" for malaria control or elimination generally implies that the investor is the public sector, and takes into account those costs and benefits which are applicable from a public welfare point of view [@Shretta_2017]; though appropriate in most cases (the government or institutions interested in public welfare primarily being the primary malaria control agents in most locations), the findings of these studies are rarely applicable to the private sector, and even less so at granular levels (such as an individual firm). In the case of a private firm not interested in "corporate social responsibility", it is not clear whether investing in malaria control would be profitable or not. This lack of clarity not only discourages investment, but also makes it difficult for governments to pinpoint the correct of amount of subsidy (if applicable) to encourage private sector scale-up in malaria control.

To address the question of the profitability of malaria control activities from the standpoint of a private firm, we analyze data during a 4 year period from a private sugar facility in Southern Mozambique. We assess the effect of indoor residual spraying (IRS) on the absenteeism and health of workers, and demonstrate that the firm's engagement in malaria control not only improved worker health, but also generated a positive return on investment from a pure accounting perspective.

The structure of the paper is as follows. After a brief review of the existing literature on the subject of private sector investment in malaria control and its effects on worker productivity, we provide an overview of the sugar company under study, and the epidemiology of malaria in the nearby area, as well as in Mozambique as a whole. We then give an overview of the data collected, and outline the theoretical and methdological assumptions and tools that underly our analysis. In the results, we show the effectiveness of the company's malaria control program in reducing absences, and translate this reduction into cost savings. Our discussion covers potential implications from this study in terms of policy and investment, as well as the paper's limitations.

@@

Regardless of the academic literature on the risk of workers becoming infected with malaria, the companies running large agricultural operations in malaria-endemic regions generally consider the disease to be an important enough threat to workers' health to merit some form of investment in malaria control [@Pluess2009]. Insecticide fumigation, bednet distribution, test-and-treat clinics, and occasional educational campaigns are common forms of corporate social responsibility to combat malaria. A World Bank study on Nigerian sugarcane cutters found that the simple availability of testing and treatment increased productivity by 10% in the weeks following the provision of services, the conclusion being that both the treatment and the test result were effective in increasing productivity, the latter simply increasing the information available which could influence personal labor allocation decisions [@Dillon2014].

Malaria may disincentivize investment in business in malaria zones. In a vicious cycle, businesses already in malaria zones may not be incentivized to invest in malaria control, perhaps because this is seen as the domain of the public sector [@Curtis2003], or perhaps due to not knowing the extent to which the benefits in health of that control will accrue to the business itself.

@@

The mill of the Maragra Açucar SA (a subsidary of the Illovo sugar company, henceforth referred to as "Maragra") was nationalized in the 1970s (like all other Mozambican mills), went through a period of low production, and then fell completely out of use by 1984. It re-opened in private hands in 1992, and was renovated by a group of international investors in 1998. Today, Maragra accounts for roughly one quarter of Mozambique's overall sugar production (second only to the nearby Xinavane mill run by the Hulett Sugar Tongaat company) [@sutton]. With a favorably close location to the port of Maputo, ample land, approximately 5,000 employees (of which three quarters are seasonal), and a mill with the capacity to process not only all the sugarcane grown on Maragra's land, but also the cane of the many nearby smallholders [@OLaughlin2016], Maragra has so far been able to weather the 2016 Mozambican crisis and concurrent collapse in global sugar prices.

>>>>

From the persepctive of the private sector, however, investing in malaria control is not so clear-cut, since the benefits are often disperse, long-term, and difficult to quantify. Public health interventions targetting malaria - and their corresponding cost-effectiveness evaluations - most often focus on impacts pertaining to public welfare, such as an increase in life years adjusted for disability or quality  [@Goodman1999] [@Shretta2016] [@Lee2017] [@Hanson2004]. Though population-level health is certainly of importance to businesses, and improvements in health incidentally improve the economy at all levels [@World1999] [@Bloom2008] [@Vecchi_2013], these improvements may be too disperse or long-term to incentivize private sector involvement in health campaigns. In other words, the returns for malaria control are less for the private sector because (i) they capture only part of the benefits and (ii) they do not benefit from the externalities.

Just as the benefits of malaria control to the private sector are more limited than those to the public sector, considerations regarding costs for a firm are also distinct than those for a government. Though many firms in endemic regions engage in malaria control programs, this should not be considered, per se, evidence of its cost-effectiveness (since the extent to which corporate social responsibility plays a role is unknown).

@@

A significant sector of the economy in Mozambique is dominated by large-scale foreign direct investment projects [@Robbins2012], and the role of the private sector in health generally, and malaria specifically, is unequivocally important. Large agriculture and extractive industry firms take up wide swaths of land and employ hundreds of thousands  [@German2013]. The Mozambican state has encouraged large-scale entreprise with the aim of general economic development [@Buur2012]. And where large firms exist, they often take on social roles such as housing and health care [@Winkler]. At times, this role is necessary from a purely practical standpoint; in other cases, it is employed under the auspices of "corporate social responsibility" [@Azemar2009] [@Curtis2003]. Regardless of the language used, it is clear that private industry plays an important role in public health in Mozambique [@Robbins2012] [@CastelBranco2014].

To address the question of the profitability of malaria control activities from the standpoint of a private firm, we analyze data during a 4 year period from a private sugar facility in Southern Mozambique. We use absenteeism as our outcome variable, assuming that it is directly correlated with the productivity losses associated with malaria infection. We assess the effect of indoor residual spraying (IRS) on absenteeism, and demonstrate that the firm's engagement in malaria control not only improved worker health, but also generated a positive return on investment from a pure accounting perspective.

The structure of the paper is as follows. We provide an overview of the sugar company under study, and the epidemiology of malaria in the nearby area, as well as in Mozambique as a whole. We then give an overview of the data collected, and outline the theoretical and methdological assumptions and tools that underly our analysis. We assess the effect of time since IRS on worker absenteeism, controlling for malaria seasonality, and segregating models for four different worker types. In the results, we show that IRS spraying is associated with a significant reduction in worker absenteeism among permanent and fieldworkers, and has little to no effect on temporary and indoor workers. We find that, in addition to reducing absence, the IRS program has a cost savings effect. Our discussion covers potential implications from this study in terms of policy and investment, as well as the paper's limitations.

## Background literature

In general, large firms operating in malaria endemic regions consider malaria to be an important enough issue to merit at least some investment [@Pluess2009]. Several studies examine the effect of foreign firms engaging in large-scale malaria control campaigns [@Han] [@Bennett_2017] [@Kaula_2017]. AngloGold-Ashanti, in partnership with local and national government in Ghana, invested in a well-rounded malaria control program in 2005, and saw worker absenteeism fall by 50% in 13 months [@anglo]. Lafarge's simulatenous investment in a comprehensive malaria control program in Benin was associated with an average 41% reduction in absenteeism among workers over the course of 4 years [@lafarge]. Zambia Sugar Plc, Zambia's largest sugar processing facility, saw annual malaria cases at its company clinic fall from nearly 3,000 in 2001 to less than 500 by 2005, following investment in a malaria control program. Marathon Oil's investment of 15 million US in vector control, education, net distribution and malaria treatment on Bioko Island in Equatorial Guinea lead to an estimated 95% reduction in the number of parasite-infected mosquitoes and 50% reduction in malaria incidence among young children [@asquino2015] [@Overgaard2012]. A PATH study in Zambia found a return on investment of 28% among three companies investing in employer-based malaria control [@Mouzin2011]. \todo{Still need to include stuff from http://gbchealth.org/wp-content/uploads/2014/03/IRS2011WorkshopReport.pdf}

Though certainly suggestive of high returns on malaria control investments, these studies generally consider population health as the outcome measure of interest, rather than worker absenteeism or productivity. Similarly, they often neglect to differentiate between those clinical costs which are absorbed by the local health system versus those which are absorbed by the firm itself. When absenteeism itself is considered, effects of malaria control have generally been found to be high, but causation is difficult to establish, given that the previous studies rely on aggregate data.

Two studies utilize worker-level data to estimate the effect of malaria control on productivity. A World Bank analysis of Nigerian sugarcane cutters found that the simple availability of testing and treatment increased productivity by 10% in the weeks following the provision of services, the conclusion being that both the treatment and the test result were effective in increasing productivity, the latter simply increasing the information available which could influence personal labor allocation decisions [@Dillon2014]. A randomized controlled trial (RCT) in Zambia showed an even greater effect from investments in preventing malaria: farmers given bed nets saw fewer days lost to illness (both directly and due to caretaking responsibilities for ill family members), translating to an increase of approximately 15% in crop yield [@Fink2015]. Though compelling, the Nigerian program only dealt with medical services (diagnostics and medication), rather than preventive interventions, and the Zambian RCT examined individual farmers, as opposed to a large firm.

In the literature, making the "investment case" for malaria control or elimination generally implies that the investor is the public sector, and takes into account those costs and benefits which are applicable from a public welfare point of view [@Shretta_2017]. For example, an economic analysis by the Corporate Alliance on Malaria in Africa on the Bioko Island Malaria Control Program found a 4:1 cost-benefit ratio, but the perspective in this case was considered to be the "community" [@lafarge]. Though appropriate in most cases to consider benefits accrued to the community (the government or institutions interested in public welfare primarily being the primary malaria control agents in most locations), the findings of these studies are rarely applicable to the private sector, and even less so at granular levels (such as an individual firm). In the case of a private firm not interested in "corporate social responsibility", it is not clear whether investing in malaria control would be profitable or not. This lack of clarity not only discourages investment, but also makes it difficult for governments to pinpoint the correct of amount of subsidy (if applicable) to encourage private sector scale-up in malaria control.

@@

@@

```{r}
# Calculate area of maragra
# x <- bairros
# x$NeighCode <- as.numeric(as.character(x$NeighCode))
# x@data <- x@data %>%
#   mutate(maragra_bairro = NeighCode %in%
#            c(601:609,701:709, 1001:1003)) %>%
#   mutate(maragra_fabrica = NeighCode %in%
#            1001:1003)
# x <- spTransform(x,CRS("+proj=utm +zone=36 +datum=WGS84 +units=m") )
# 
# the_bairro <- x[x$maragra_bairro,]
# the_fabrica <- x[x$maragra_fabrica,]
# # the_bairro <- spTransform(the_bairro, CRS("+proj=utm +zone=36 +datum=WGS84 +units=m") )
# # the_fabrica <- spTransform(the_fabrica, CRS("+proj=utm +zone=36 +datum=WGS84 +units=m") )
# 
# rgeos::gArea(the_bairro) / 1000000
# rgeos::gArea(the_fabrica) / 1000000

# Used this for plantation estimation: https://www.daftlogic.com/projects-google-maps-area-calculator-tool.htm
```

The mill of the Maragra Açucar SA (a subsidary of the Illovo sugar company, henceforth referred to as "Maragra") was nationalized in the 1970s (like all other Mozambican mills), went through a period of low production, and then fell completely out of use by 1984. It re-opened in private hands in 1992, and was renovated by a group of international investors in 1998. Today, Maragra accounts for roughly one quarter of Mozambique's overall sugar production (second only to the nearby Xinavane mill run by the Hulett Sugar Tongaat company) [@sutton]. With a favorably close location to the port of Maputo, ample land (approximately 90 squared kilometers of plantation, and 5 squared kilometers of factory grounds), approximately 5,000 employees (of which three quarters are seasonal), and a mill with the capacity to process not only all the sugarcane grown on Maragra's land, but also the cane of the many nearby smallholders [@OLaughlin2016], Maragra has so far been able to weather the 2016 Mozambican crisis and concurrent collapse in global sugar prices.

### 2018-04-20 inst/rmd/02_methods.Rmd
major overhaul

Maragra regularly employs IRS at on-site worker households in order to reduce those workers' (and their families') risk of malaria infection. Workers living off-site (our control group) also may have received IRS at some point during the study period (from government programs). Even though we do not have reliable person-level data on IRS carried out by the government, off-site workers are a suitable control in the sense that they represent "business as usual" (ie, what would happen if the company carried out no IRS and relied solely on public interventions). Using company HR and clinical records, we were able to identify absences and episodes of clinical malaria among all workers, as well as identify the time since the most recent IRS episode before the onsent of absence or illness.

@@

We sought to understand the effect of IRS on individual workers' likelihood of absence from work as well as their likelihood of clinical malaria. To estimate this effect, we estimated separate models for absence and illness. We employed linear mixed effects models [@Minalu2011] [@Liu2010] (with individual worker fixed effects, and random effects for other covariates)  in lieu of interrupted time series [@Lopez_Bernal_2016] so as to explicitly adjust for confounders [@Bell2014]. We divide into 4 different models for different worker types, so as to account for the potential time-varying effects of worker type on risk of malaria. Our econometric model is relatively straightforward:

$$
\hat{Y_{it}} = \beta_{0} +  (\beta_{1}) (\text{Season}_{it}) + (\beta_2{IRS}*\beta_3{IRS_it}) + ... + \epsilon_{it}
$$

$\hat{Y}$ is the rate of absence. $\beta_{1}$ represents the clinical malaria incidence at that time in the entire district of Manhiça.  Our demographic confounders (represented by $...$) are sex, age, and worker department (field, factory, or administrative). Our intervention whether the residence of the worker in question was treated in the last year, and, if so, the time since treatment (represented above as the interaction term, where where $_t$ represents time elapsed since commencement of the most recent IRS campaign). We define the malaria season as any time during which the clinical incidence of malaria in the district of Manhiça was at or greater than the median clinical incidence of malaria for the entire study period. These weeks are flagged as red in Figure 1, Panel A.  By using clinical incidence of the area of residence of the workers (as opposed to more typical proxies for malaria risk, such as only rainy vs. non rainy season), our seasonality estimate is a closer approximation of true malaria risk, incorporating lagged effects such as the incubation period of the parasite, as well as any inherent non-linear effects of weather. In addition, we adjust for daily precipitation

@@

We collected absenteeism and demographic data data on `r n` workers from 2012 through 2016. Workers were approximately 60% male, and more than 80% fieldworkers (predominantly cane-cutters). Most were in their 20s and 30s and employed on temporary contracts. Table 1 shows overall worker details.

```{r, results='asis'}
x <- workers %>%
  filter(oracle_number %in% unique(ab_panel$oracle_number))
x$age <- round(as.numeric(as.Date('2014-01-01') - x$date_of_birth) / 365.25, digits = -1)
x$status <- x$permanent_or_temporary

calculate <- function(var= 'sex',
                      data){
  out <- 
    data %>%
    group_by_(var) %>%
    summarise(n = n()) %>%
    ungroup() %>%
    mutate(p = round(n / sum(n) * 100, digits = 2)) %>%
    mutate(variable = '') 
    names(out)[1] <- ' '
    out$variable[1] <- Hmisc::capitalize(var)

    out <- out %>% dplyr::select(variable, ` `, n, p)
    out$` ` <- as.character(out$` `)
  return(out)
}
out_list <- list()
vars <- c('department', 
           'sex',
           'status',
           'age')
for(i in 1:length(vars)){
  message(i)
  y <- calculate(var = vars[i], data = x)
  out_list[[i]] <- y
}
demo_table <- bind_rows(out_list)
names(demo_table) <- c('Variable', ' ', 'Workers', '(%)')
print(xtable::xtable(demo_table,
                     caption = 'Overall worker characteristics',
                     align = c('r', 'r', rep('l', 3))),
      include.rownames = FALSE)

ov_ab <- round(100 * mean(ab_panel$absent[between(ab_panel$date, as.Date('2012-01-01'), as.Date('2016-12-31'))]), digits = 2)

ov_cl <- length(which(between(clinic$date, as.Date('2012-01-01'), as.Date('2016-12-31'))))
```

The overall absenteeism rate during the study period was 5.56% among all workers. The number of malaria cases registered at the company clinic during that time was 1873, for an approximate annualized clinical incidence of 110 cases per 1,000.

@@

```{r, fig.width = 3, fig.height = 2.5, fig.align='center', fig.cap = 'Age distribution by sex and worker status'}
x$age <- round(as.numeric(as.Date('2014-01-01') - x$date_of_birth) / 365.25, digits = 2)
g1 <- ggplot(data = x,
       aes(x = age,
           group = sex,
           fill = sex)) +
  geom_density(alpha = 0.8) +
  facet_wrap(~permanent_or_temporary) +
  theme_maragra() +
  labs(x = 'Age',
       y = 'Density') +
  scale_fill_manual(name = 'Sex',
                    values = c('lightblue', 'grey'))
g1
```

>>>>

Maragra regularly employs IRS at on-site worker households in order to reduce those workers' (and their families') risk of malaria infection. IRS works by killing the malaria vector (mosquito), thereby preventing infection of the household occupants. When administered correctly, IRS is a low-risk intervention to its recipients, and is assumed not to affect absenteeism in the short-term (to the extent that it may cause negative side-effects, these are generally long-term). It is preventive only, and does not cure current malaria infection, nor does it affect the parasite load of mosquitoes. Workers living off-site (our control group) also may have received IRS at some point during the study period (from government programs). Even though we do not have reliable person-level data on IRS carried out by the government, off-site workers are a suitable control in the sense that they represent "business as usual" (ie, what would happen if the company carried out no IRS and relied solely on public interventions). Using company HR and clinical records, we were able to identify absences and episodes of clinical malaria among all workers, as well as identify the time since the most recent IRS episode before the onsent of absence or illness.

@@

We sought to understand the effect of IRS on individual workers' likelihood of absence from work as well as their likelihood of clinical malaria. To estimate this effect, we estimated separate models for absence and illness. We employed linear mixed effects models [@Minalu2011] [@Liu2010] (with individual worker fixed effects, and random effects for other covariates)  in lieu of interrupted time series [@Lopez_Bernal_2016] so as to explicitly adjust for confounders [@Bell2014]. We divide into 4 different models for different worker types, so as to account for the potential time-varying effects of worker type on risk of malaria. Our model specification is as follows

$$
\hat{Y_{it}} = \hat{\beta}_{0} +  \hat{\beta}_{1}\text{Season}_{t} * (\hat{\beta}_2{IRS_{it}}*\hat{\beta}_3{IRStime_{it}}) + \alpha_i + \delta_t + \upsilon_{it}
$$

$\hat{Y}_{it}$ is the rate of absence. $\beta_{1}$ is the binary "season" variable, imputed from overall district clinical incidence. Our intervention is whether the residence of the worker in question was treated in the last year, and, if so, the time since treatment, represented, respectively, by $\beta_{2}$ and $\beta_{3}$. $\alpha_i$ reprsents the time invariant worker fixed effects, and $\delta_i$ represents the fixed effect of the particular malaria season. $\upsilon$ is the error term.

We define the malaria season as any time during which the clinical incidence of malaria in the district of Manhiça was at or greater than the median clinical incidence of malaria for the entire study period. These weeks are flagged as red in Figure 1, Panel A.  By using clinical incidence of the area of residence of the workers (as opposed to more typical proxies for malaria risk, such as only rainy vs. non rainy season), our seasonality estimate is a closer approximation of true malaria risk, incorporating lagged effects such as the incubation period of the parasite, as well as any inherent non-linear effects of weather. In addition, we adjust for daily precipitation; though its lagged effect on malaria incidence is likely captured by the seasonality term, we include rainfall since it's immediate effect (through its impact on transportation and working conditions) may also affect absenteeism.

@@

We collected absenteeism and demographic data data on `r n` workers from 2012 through 2016. Workers were approximately 60% male, and more than 80% fieldworkers (predominantly cane-cutters). Most were in their 20s and 30s and employed on temporary contracts. Table 1 shows overall worker details divided by "Treatment" versus "Control" status. "Treatment" is considered the time beginning at IRS administration and ending one year later; "Control" is considered time observed either prior to IRS administration (and greater than 1 year after previous IRS administration) or among workers who never received IRS. Because of the temporal nature of these categories, many workers belong to both treatment and control groups, albeit at different times.

```{r, results='asis'}
# x <- ab_panel %>%
#   left_join(workers %>%
#               dplyr::select(oracle_number, date_of_birth, permanent_or_temporary, sex))
x <- model_data
x$age <- round(as.numeric(as.Date('2014-01-01') - x$date_of_birth) / 365.25, digits = -1)
x$status <- x$permanent_or_temporary

calculate <- function(var= 'sex',
                      data){
  out <- 
    data %>%
    group_by_(var) %>%
    summarise(`Treatment workers` = length(unique(oracle_number[!days_since %in% c('Never', 'Before')])),
              `Treatment days` = length(oracle_number[!days_since %in% c('Never', 'Before')]),
              `Control workers` = length(unique(oracle_number[days_since %in% c('Never', 'Before')])),
              `Control days` = length(oracle_number[days_since %in% c('Never', 'Before')])) %>%
    ungroup() %>%
    # mutate(p = round(n / sum(n) * 100, digits = 2)) %>%
    mutate(variable = '') 
    names(out)[1] <- ' '
    out$variable[1] <- Hmisc::capitalize(var)

    out <- out %>% dplyr::select(variable, ` `, `Treatment workers`, `Treatment days`, `Control workers`, `Control days`)
    out$` ` <- as.character(out$` `)
  return(out)
}
out_list <- list()
vars <- c('department', 
           'sex',
           'status',
           'age')
for(i in 1:length(vars)){
  message(i)
  y <- calculate(var = vars[i], data = x)
  out_list[[i]] <- y
}
demo_table <- bind_rows(out_list)
names(demo_table)[1:2] <- c('Variable', ' ')
print(xtable::xtable(demo_table,
                     caption = 'Overall worker characteristics',
                     align = c('r', 'r', rep('l', 5))),
      include.rownames = FALSE)

```

@@

```{r, fig.width = 5, fig.height = 3.5, fig.align='center', fig.cap = 'Age distribution by sex and worker status'}
x$age <- round(as.numeric(as.Date('2014-01-01') - x$date_of_birth) / 365.25, digits = 2)
g1 <- ggplot(data = x,
       aes(x = age,
           group = sex,
           fill = sex)) +
  geom_density(alpha = 0.8) +
  facet_wrap(~permanent_or_temporary) +
  theme_maragra() +
  labs(x = 'Age',
       y = 'Density') +
  scale_fill_manual(name = 'Sex',
                    values = c('lightblue', 'grey'))
g1
```
